KiCS2: A New Compiler from Curry to Haskell

### 2011-05-06 papers/wflp2011/paper.tex

  xorSelf x = xor x x
\end{curry}
%
and the expression \ccode{xorSelf aBool}.
If we interpret this program as a term rewriting system,
we could have the reduction
\begin{lstlisting}
  xorSelf aBool  $\to~$  xor aBool aBool     $\to~$  xor True aBool
                 $\to~$  xor True False      $\to~$  True
\end{lstlisting}
leading to the unintended result \code{True}.
Note that this result cannot be obtained if we use a strict strategy
where arguments are evaluated before the function calls.
In order to avoid dependencies on the evaluation strategies
and exclude such unintended results,
Gonz\'alez-Moreno et al.\ \cite{GonzalezEtAl99} proposed
the rewriting logic CRWL as a logical
(execution- and strategy-independent) foundation for declarative
programming with non-strict and non-deterministic operations.  This
logic specifies the \emph{call-time choice} semantics \cite{Hussmann92}
\label{ctc-semantics}
where values of the arguments of an operation are determined before the
operation is evaluated. Note that this does not necessarily require
an eager evaluation of arguments.
Actually, \cite{AlbertHanusHuchOliverVidal05,LopezRodriguezSanchez07}
define lazy evaluation strategies for functional logic programs
with call-time choice semantics where actual arguments passed to
operations are shared. Hence, we can evaluate the expression above
lazily, provided that all occurrences of \code{aBool}
are shared so that all of them reduce either to \code{True} or to \code{False}.
The requirements of this call-time choice semantics are the
reason why it is not simply possible to use list comprehensions
or non-determinism monads for a straightforward implementation
of functional logic programs in Haskell \cite{FischerKiselyovShan09}.

@@

\label{sec:no-logic-vars}
First of all, we assume that our programs
\emph{do not contain logic variables}.
This assumption can be made since it has been shown
\cite{AntoyHanus06ICLP} that logic variables can be replaced
by non-deterministic ``generators,''
i.e., operations that evaluate to all possible values of the type of
the logic variable. For instance, a Boolean logic variable
can be replaced by the generator \code{aBool} defined above.

@@

\begin{figure}[t]
\[
\begin{array}{@{}rcl@{~~~~~~~~}l}
  e & ::= & x & \mbox{$x$ is a variable}\\
    &  |  & c(e_1,\ldots,e_n) & \mbox{$c$ is an $n$-ary constructor symbol}\\
    &  |  & f(e_1,\ldots,e_n) & \mbox{$f$ is an $n$-ary function symbol}\\
    &  |  & \Choice{e_1}{e_2} & \mbox{choice} \\[1ex]
  D & ::= & f(x_1,\ldots,x_n) = e
                & \mbox{$n$-ary function $f$ with a single rule}\\
    &  |  & f(c(y_1,\ldots,y_m),x_2,\ldots,x_n) = e
                & \mbox{matching rule for $n$-ary function $f$}\\
    &     & & \mbox{$c$ is an $m$-ary constructor symbol}\\
  P & ::= & D_1 \ldots D_k
\end{array}
\]
\caption{Uniform programs ($e$: expressions, $D$: definitions, $P$: programs)}
\label{fig:uniform}
\end{figure}
%
Finally, we assume that the pattern matching strategy
is explicitly encoded in individual matching functions.
In contrast to \cite{AlbertHanusHuchOliverVidal05},
where the pattern matching strategy is encoded in case expressions,
we assume that each case expression is transformed into
a new operation in order to avoid complications by translating
nested case expressions.
Thus, we assume that all programs are \emph{uniform}
according to the definition in Fig.~\ref{fig:uniform}.\footnote{%
A variant of uniform programs have been considered in
\cite{MorenoEtAl90ALP} to define lazy narrowing strategies
for functional logic programs. Although the motivations are similar,
our notion of uniform program is more restrictive since we
allow only a single non-variable argument in each left-hand side
of a rule.
Uniform programs have been also applied in
\cite{Zartmann97SAS} to define a denotational analysis
of functional logic programs.}
There, the variables in the left-hand sides of each rule are pairwise
different, and
the constructors in the left-hand sides of the matching rules of each function
are pairwise different.
Uniform programs have a simple form of pattern matching:
either a function is defined by a single rule without pattern matching,
or it is defined by rules with only one constructor in the left-hand
side of each rule, and in the same argument for all rules.%
\footnote{For simplicity, we require in
Fig.~\ref{fig:uniform} that the matching argument is always the
first one, but one can also choose any other argument.}
For instance, the operation \code{xor} defined above can be transformed
into the following uniform program:
%
\begin{curry}
  xor True   x = xor' x
  xor False  x = x
  xor' False = True
  xor' True  = False
\end{curry}
%
In particular, there are no overlapping rules for functions
(except for the choice operation \ccode{?} which is considered as predefined).
Antoy \cite{Antoy01PPDP} showed that each functional logic program,
i.e., each constructor-based conditional term rewriting system,
can be translated into an equivalent unconditional term rewriting system
without overlapping rules but containing choices in the right-hand sides,
also called LOIS (limited overlapping inductively sequential) system.
Furthermore, Brassel \cite{Brassel11Thesis} showed the semantical
equivalence of narrowing computations in LOIS systems
and rewriting computations in uniform programs.
Due to these results, uniform programs are a reasonable intermediate
language for our translation into Haskell which will be presented
in the following.

@@

As mentioned above, our implementation is based on the explicit
representation of non-deterministic results in a data structure.
This can easily be achieved by adding a constructor
to each data type to represent a choice between two values.
For instance, one can redefine the data type for Boolean values
as follows:
\begin{haskell}
  data Bool = False | True | Choice Bool Bool
\end{haskell}
Thus, we can implement the non-deterministic operation \code{aBool}
defined in Section~\ref{ex:aBool} as:
\begin{haskell}
  aBool = Choice True False
\end{haskell}
If operations can deliver non-deterministic values,
we have to extend the rules for operations defined by
pattern matching so that they do not fail on non-deterministic argument
values. Instead, they move the non-deterministic choice one level above,
i.e., a choice in some argument leads to a choice in any result
of this operation (this is also called a ``pull-tab'' step in
\cite{AlqaddoumiAntoyFischerReck10}). For instance,
the rules of the uniform operation \code{xor} shown above
are extended as follows:
%
\begin{curry}
  xor True           x = xor' x
  xor False          x = x
  xor (Choice x1 x2) x = Choice (xor x1 x) (xor x2 x)

  xor' False          = True
  xor' True           = False
  xor' (Choice x1 x2) = Choice (xor' x1) (xor' x2)
\end{curry}
%
The operation \code{xorSelf} is not defined by a pattern matching rule
and, thus, need not be changed.
If we evaluate the expression \ccode{xorSelf aBool}, we get the result
\begin{lstlisting}
  Choice (Choice False True) (Choice True False)
\end{lstlisting}
How can we interpret this result?
In principle, the choices represent different possible values.
Thus, if we want to show the different values of an expression
(which is usually the task of a top-level ``read-eval-print'' loop),
we enumerate all values contained in the choices.
These are \code{False}, \code{True}, \code{True}, and \code{False}
in the result above.
Unfortunately, this does not conform to the call-time choice semantics
discussed in Section~\ref{ctc-semantics} which should not yield
a value like \code{True}.
The call-time choice semantics requires that the choice
of a value made for
the initial expression \code{aBool} should be consistent
in the entire computation.
For instance, if we select the value \code{False} for the
expression \code{aBool}, this selection should be made
at all other places where this expression might have been copied
during the computation. However, our initial implementation
duplicates the initially single \code{Choice} into finally three
occurrences of \code{Choice}.

We can correct this unintended behavior of our implementation
by identifying different \code{Choice} occurrences that are duplicates
of some single \code{Choice}. This can be easily done by attaching
a unique identifier, e.g., a number, to each choice:
\begin{haskell}
  type ID = Integer
  data Bool = False | True | Choice ID Bool Bool
\end{haskell}
Furthermore, we modify the \code{Choice} pattern rules so that
the identifiers will be kept, e.g.,
\begin{curry}
  xor (Choice i x1 x2) x = Choice i (xor x1 x) (xor x2 x)
\end{curry}
If we evaluate the expression \ccode{xorSelf aBool} and
assign the number \code{1}
to the choice of \code{aBool}, we obtain the result
\begin{lstlisting}
  Choice 1 (Choice 1 False True) (Choice 1 True False)
\end{lstlisting}
When we show the values contained in this result,
we have to make \emph{consistent} selections in choices with
same identifiers. This, if we select the left branch as the value
of the outermost \code{Choice}, we also have to select the left branch
in the selected argument \code{(Choice 1 False True)} so that only
the value \code{False} is possible here.
Similarly, if we select the right branch as the value of the outermost
\code{Choice}, we also have to select the right branch in
its selected argument \code{(Choice 1 True False)} which yields the only
value \code{False}.

Note that each \code{Choice} occurring for the first time in a computation
has to get its own unique identifier.
For instance, if we evaluate the expression \ccode{xor aBool aBool},
the two occurrences of \code{aBool} assign different identifiers
to their \code{Choice} constructor (e.g., \code{1} for the left
and \code{2} for the right \code{aBool} argument) so that this evaluates to
\begin{lstlisting}
  Choice 1 (Choice 2 False True) (Choice 2 True False)
\end{lstlisting}
Here we can make different selections for the outer and inner
\code{Choice} constructors so that this non-deterministic result represents
four values.

To summarize, our implementation is based on the following principles:
\begin{enumerate}
\item Each non-deterministic choice is represented by a \code{Choice}
constructor with a unique identifier.
\item When matching a \code{Choice} constructor, the choice is
moved to the result of this operation with the same identifier,
i.e., a non-deterministic argument
yields non-deterministic results for each of the argument's values.
\item Each choice introduced in a computation gets its own unique
identifier.
\end{enumerate}
The latter principle requires the creation of fresh identifiers
during a computation---a non-trivial problem in functional languages.
One possibility is the use of a global counter
that is accessed by unsafe features whenever a new identifier
is required. Unfortunately, unsafe features inhibit the use
of optimization techniques developed for purely functional programs
and make the application of advanced evaluation and search
strategies (e.g., parallel strategies) more complex.
Therefore, we avoid unsafe features in our implementation.
Instead, we thread some global information
through our program in order to supply fresh references
at any point of a computation.
For this purpose, we assume a type \code{IDSupply}
with operations
\begin{haskell}
  initSupply  :: IO IDSupply
  thisID      :: IDSupply -> ID
  leftSupply  :: IDSupply -> IDSupply
  rightSupply :: IDSupply -> IDSupply
\end{haskell}
and add a new argument of type \code{IDSupply} to each operation
of the source program, i.e., a Curry operation of type
\begin{curry}
  f :: $\tau_1$ ->$~\cdots~$->$~\tau_n~$->$~\tau$
\end{curry}
is translated into a Haskell function of type
\begin{haskell}
  f :: $\tau_1$ ->$~\cdots~$->$~\tau_n~$-> IDSupply ->$~\tau$
\end{haskell}
%
Conceptually, one can consider \code{IDSupply}
as an infinite set of identifiers that is created at the beginning
of an evaluation by the operation \code{initSupply}.
The operation \code{thisID} takes some identifier from this set, and
\code{leftSupply} and \code{rightSupply} splits this set
into two disjoint subsets without the identifier
obtained by \code{thisID}.
The split operations \code{leftSupply} and \code{rightSupply}
are used when an operation calls two\footnote{The
extension to more than two is straightforward.}
other operations in the right-hand side of a rule. In this case,
the other operations must be supplied with their individual disjoint
identifier supplies. For instance, the operation \code{main}
defined by
\label{sec:xor-main}
\begin{curry}
  main :: Bool
  main = xorSelf aBool
\end{curry}
is translated into
\begin{haskell}
  main :: IDSupply -> Bool
  main s = xorSelf (aBool (leftSupply s)) (rightSupply s)
\end{haskell}
Any choice in the right-hand side of a rule gets its own identifier
by the operation \code{thisID}, as in
\begin{haskell}
  aBool s = Choice (thisID s) True False
\end{haskell}
There are various implementations of \code{IDSupply}.
The simplest implementation uses unbounded integers:
\begin{haskell}
  type IDSupply = Integer
  initSupply    = return 1
  thisID      n = n
  leftSupply  n = 2*n
  rightSupply n = 2*n+1
\end{haskell}
There are other more sophisticated implementations available
\cite{AugustssonRittriSynek94}.
Actually, our compilation system is parameterized over different
implementations of \code{IDSupply}
in order to perform some experiments and choose
the most appropiate for a given application.
Each implementation must ensure that, if $s$ is a value of type
\code{IDSupply}, then \code{thisID($o_1$($\ldots$($o_n$ $s$)$\ldots$))}
and \code{thisID($o'_1$($\ldots$($o'_m$ $s$)$\ldots$))}
are different identifiers provided that
$o_i,o'_j \in \{\code{leftSupply},\code{rightSupply}\}$
and $o_1\cdots o_n \neq o'_1\cdots o'_m$.

@@

So far, our generated operations compute all the non-deterministic
values of an expression represented by a structure containing
\code{Choice} constructors. In order to extract the various
values from this structure, we have to define operations
that compute all possible choices in some order where the choice identifiers
are taken into account.
To provide a common interface for such operations, we introduce
a data type to represent the general outcome of a computation,
\begin{haskell}
  data Try a = Val a | Choice ID a a | Fail
\end{haskell}
together with an auxiliary operation:\footnote{Note that the
operation \code{try} is not really polymorphic but overloaded
for each data type and, therefore, defined in instances of some type class.}
\begin{haskell}
  try :: a -> Try a
  try (Choice i x y) = Choice i x y
  try Fail           = Fail
  try x              = Val x
\end{haskell}
In order to take the identity of choices into account when extracting values,
one has to remember which choice (e.g., left or right branch)
has been made for some particular choice.
Therefore, we introduce the type
\begin{haskell}
  data Choice = NoChoice | ChooseLeft | ChooseRight
\end{haskell}
where \code{NoChoice} represents the fact that a choice has not yet been made.
Furthermore, we need operations to lookup the current choice
for a given identifier or change its choice:
\begin{haskell}
  lookupChoice :: ID -> IO Choice
  setChoice    :: ID -> Choice -> IO ()
\end{haskell}
In Haskell, there are different possibilities to implement a mapping
from choice identifiers to some value of type \code{Choice}.
Our implementation supports various options together with different
implementations of \code{IDSupply}.
For instance, a simple but efficient implementation can be obtained by
using updatable values, i.e., the Haskell type \code{IORef}.
In this case, choice identifiers are memory cells instead of integers:
\begin{haskell}
  newtype ID = ID (IORef Choice)
\end{haskell}
Consequently, an implementation of \code{IDSupply} requires
an infinite set of memory cells which can represented as a tree structure:
\begin{haskell}
  data IDSupply = IDSupply ID IDSupply IDSupply
  thisID      (IDSupply r _ _) = r
  leftSupply  (IDSupply _ s _) = s
  rightSupply (IDSupply _ _ s) = s
  \end{haskell}
The infinite tree of memory cells (with initial value \code{NoChoice})
can be constructed as follows,
where \code{unsafeInterleaveIO} is used to construct
the tree on demand:
\begin{haskell}
  initSupply = getIDTree$\listline$
  getIDTree = do s1 <- unsafeInterleaveIO getIDTree
                 s2 <- unsafeInterleaveIO getIDTree
                 r  <- unsafeInterleaveIO (newIORef NoChoice)
                 return (IDSupply (ID r) s1 s2)
  \end{haskell}
Using memory cells,
the implementation of the lookup and set operations is straightforward:
\begin{haskell}
  lookupChoice (ID ref) = readIORef ref
  setChoice (ID ref) c = writeIORef ref c
\end{haskell}
Now we can print all values contained in a choice structure
in a depth-first manner by the following operation:
\label{sec:printValsDFS}
\begin{haskell}
  printValsDFS :: Try a -> IO ()
  printValsDFS (Val v)          = print v
  printValsDFS Fail             = return ()
  printValsDFS (Choice i x1 x2) = lookupChoice i >>= choose
   where
    choose ChooseLeft  = printValsDFS (try x1)
    choose ChooseRight = printValsDFS (try x2)
    choose NoChoice    = do newChoice ChooseLeft  x1
                            newChoice ChooseRight x2

@@

Instead of choosing different search strategies to evaluate
top-level expressions, our system also contains a primitive operation
\begin{curry}
 searchTree :: a -> IO (SearchTree a)
\end{curry}
to translate the search space caused by the evaluation of its argument
into a tree structure of the form
\begin{curry}
 data SearchTree a = Value a | Fail | Or (SearchTree a) (SearchTree a)
\end{curry}
With this primitive,
the programmer can define its own search strategy or
collect all non-deterministic values into a list structure
for further processing \cite{BrasselHuch07}.

@@

We have already discussed in the introduction the various
efforts to implement functional logic languages,
like the construction of abstract machines
\cite{AntoyHanusLiuTolmach05,HanusSadre99JFLP,Lux99FLOPS} and
the compilation into Prolog \cite{AntoyHanus00FROCOS}
or Haskell \cite{BrasselFischerHanusReck11,BrasselHuch07,BrasselHuch09}.
Our benchmarks show that an efficient implementation
by compiling into
a functional language depends on carefully handling
the sharing of non-deterministic choices. For instance, our previous
implementation \cite{BrasselFischerHanusReck11}, where sharing
is explicitly managed by the monadic techniques proposed
in \cite{FischerKiselyovShan09}, has not satisfied the expectations
that came from the benchmarks reported in \cite{FischerKiselyovShan09}.
Due to these experiences, we used in our new compiler the
compilation scheme initially proposed in \cite{BrasselFischer08IFL}
which produces much faster code, as shown in our benchmarks.

>>>>

  xorSelf x = xor x x
\end{curry}
%
and the expression \ccode{xorSelf aBool}.
If we interpret this program as a term rewriting system,
we could have the reduction
\begin{haskell}
  xorSelf aBool  $\to~$  xor aBool aBool     $\to~$  xor True aBool
                 $\to~$  xor True False      $\to~$  True
\end{haskell}
leading to the unintended result \code{True}.
Note that this result cannot be obtained if we use a strict strategy
where arguments are evaluated before the function calls.
In order to avoid dependencies on the evaluation strategies
and exclude such unintended results,
Gonz\'alez-Moreno et al.\ \cite{GonzalezEtAl99} proposed
the rewriting logic CRWL as a logical
(execution- and strategy-independent) foundation for declarative
programming with non-strict and non-deterministic operations.  This
logic specifies the \emph{call-time choice} semantics \cite{Hussmann92}
\label{ctc-semantics}
where values of the arguments of an operation are determined before the
operation is evaluated. Note that this does not necessarily require
an eager evaluation of arguments.
Actually, \cite{AlbertHanusHuchOliverVidal05,LopezRodriguezSanchez07}
define lazy evaluation strategies for functional logic programs
with call-time choice semantics where actual arguments passed to
operations are shared. Hence, we can evaluate the expression above
lazily, provided that all occurrences of \code{aBool}
are shared so that all of them reduce either to \code{True} or to \code{False}.
The requirements of this call-time choice semantics are the
reason why it is not simply possible to use list comprehensions
or non-determinism monads for a straightforward implementation
of functional logic programs in Haskell \cite{FischerKiselyovShan09}.

@@

\label{sec:no-logic-vars}
First of all, we assume that our programs
\emph{do not contain logic variables}.
This assumption can be made since it has been shown
\cite{AntoyHanus06ICLP} that logic variables can be replaced
by non-deterministic ``generators'',
i.e., operations that evaluate to all possible values of the type of
the logic variable. For instance, a Boolean logic variable
can be replaced by the generator \code{aBool} defined above.

@@

\begin{figure}[t]
\[
\begin{array}{@{}rcl@{~~~~~~~~}l}
  e & ::= & x & \mbox{$x$ is a variable}\\
    &  |  & c(e_1,\ldots,e_n) & \mbox{$c$ is an $n$-ary constructor symbol}\\
    &  |  & f(e_1,\ldots,e_n) & \mbox{$f$ is an $n$-ary function symbol}\\
    &  |  & \Choice{e_1}{e_2} & \mbox{choice} \\[1ex]
  D & ::= & f(x_1,\ldots,x_n) = e
                & \mbox{$n$-ary function $f$ with a single rule}\\
    &  |  & f(c(y_1,\ldots,y_m),x_2,\ldots,x_n) = e
                & \mbox{matching rule for $n$-ary function $f$}\\
    &     & & \mbox{$c$ is an $m$-ary constructor symbol}\\
  P & ::= & D_1 \ldots D_k
\end{array}
\]
\caption{Uniform programs ($e$: expressions, $D$: definitions, $P$: programs)}
\label{fig:uniform}
\end{figure}
%
Finally, we assume that the pattern matching strategy
is explicitly encoded in individual matching functions.
In contrast to \cite{AlbertHanusHuchOliverVidal05},
where the pattern matching strategy is encoded in case expressions,
we assume that each case expression is transformed into
a new operation in order to avoid complications by translating
nested case expressions.
Thus, we assume that all programs are \emph{uniform}
according to the definition in Fig.~\ref{fig:uniform}.\footnote{%
A variant of uniform programs has been considered in
\cite{MorenoEtAl90ALP} to define lazy narrowing strategies
for functional logic programs. Although the motivations are similar,
our notion of uniform programs is more restrictive since we
allow only a single non-variable argument in each left-hand side
of a rule.
Uniform programs have also been applied in
\cite{Zartmann97SAS} to define a denotational analysis
of functional logic programs.}
There, the variables in the left-hand sides of each rule are pairwise
different, and
the constructors in the left-hand sides of the matching rules of each function
are pairwise different.
Uniform programs have a simple form of pattern matching:
either a function is defined by a single rule without pattern matching,
or it is defined by rules with only one constructor in the left-hand
side of each rule, and in the same argument for all rules.%
\footnote{For simplicity, we require in
Fig.~\ref{fig:uniform} that the matching argument is always the
first one, but one can also choose any other argument.}
For instance, the operation \code{xor} defined above can be transformed
into the following uniform program:
%
\begin{curry}
  xor True   x = xor' x
  xor False  x = x
  xor' False = True
  xor' True  = False
\end{curry}
%
In particular, there are no overlapping rules for functions
(except for the choice operation \ccode{?} which is considered as predefined).
Antoy \cite{Antoy01PPDP} showed that each functional logic program,
i.e., each constructor-based conditional term rewriting system,
can be translated into an equivalent unconditional term rewriting system
without overlapping rules but containing choices in the right-hand sides,
also called LOIS (limited overlapping inductively sequential) system.
Furthermore, Bra{\ss}el \cite{Brassel11Thesis} showed the semantical
equivalence of narrowing computations in LOIS systems
and rewriting computations in uniform programs.
Due to these results, uniform programs are a reasonable intermediate
language for our translation into Haskell which will be presented
in the following.

@@

As mentioned above, our implementation is based on the explicit
representation of non-deterministic results in a data structure.
This can easily be achieved by adding a constructor
to each data type to represent a choice between two values.
For instance, one can redefine the data type for Boolean values
as follows:
\begin{haskell}
  data Bool = False | True | Choice Bool Bool
\end{haskell}
Thus, we can implement the non-deterministic operation \code{aBool}
defined in Section~\ref{ex:aBool} as:
\begin{haskell}
  aBool = Choice True False
\end{haskell}
If operations can deliver non-deterministic values,
we have to extend the rules for operations defined by
pattern matching so that they do not fail on non-deterministic argument
values. Instead, they move the non-deterministic choice one level above,
i.e., a choice in some argument leads to a choice in any result
of this operation (this is also called a ``pull-tab'' step in
\cite{AlqaddoumiAntoyFischerReck10}). For instance,
the rules of the uniform operation \code{xor} shown above
are extended as follows:
%
\begin{haskell}
  xor True           x = xor' x
  xor False          x = x
  xor (Choice x1 x2) x = Choice (xor x1 x) (xor x2 x)

  xor' False          = True
  xor' True           = False
  xor' (Choice x1 x2) = Choice (xor' x1) (xor' x2)
\end{haskell}
%
The operation \code{xorSelf} is not defined by a pattern matching rule
and, thus, need not be changed.
If we evaluate the expression \ccode{xorSelf aBool}, we get the result
\begin{haskell}
  Choice (Choice False True) (Choice True False)
\end{haskell}
How can we interpret this result?
In principle, the choices represent different possible values.
Thus, if we want to show the different values of an expression
(which is usually the task of a top-level ``read-eval-print'' loop),
we enumerate all values contained in the choices.
These are \code{False}, \code{True}, \code{True}, and \code{False}
in the result above.
Unfortunately, this does not conform to the call-time choice semantics
discussed in Section~\ref{ctc-semantics} which excludes
a value like \code{True}.
The call-time choice semantics requires that the choice
of a value made for
the initial expression \code{aBool} should be consistent
in the entire computation.
For instance, if we select the value \code{False} for the
expression \code{aBool}, this selection should be made
at all other places where this expression might have been copied
during the computation. However, our initial implementation
duplicates the initially single \code{Choice} into finally three
occurrences of \code{Choice}.

We can correct this unintended behavior of our implementation
by identifying different \code{Choice} occurrences that are duplicates
of some single \code{Choice}. This can be easily done by attaching
a unique identifier, e.g., a number, to each choice:
\begin{haskell}
  type ID = Integer
  data Bool = False | True | Choice ID Bool Bool
\end{haskell}
Furthermore, we modify the \code{Choice} pattern rules so that
the identifiers will be kept, e.g.,
\begin{haskell}
  xor (Choice i x1 x2) x = Choice i (xor x1 x) (xor x2 x)
\end{haskell}
If we evaluate the expression \ccode{xorSelf aBool} and
assign the number \code{1}
to the choice of \code{aBool}, we obtain the result
\begin{haskell}
  Choice 1 (Choice 1 False True) (Choice 1 True False)
\end{haskell}
When we show the values contained in this result,
we have to make \emph{consistent} selections in choices with
same identifiers. This, if we select the left branch as the value
of the outermost \code{Choice}, we also have to select the left branch
in the selected argument \code{(Choice 1 False True)} so that only
the value \code{False} is possible here.
Similarly, if we select the right branch as the value of the outermost
\code{Choice}, we also have to select the right branch in
its selected argument \code{(Choice 1 True False)} which yields the only
value \code{False}.

Note that each \code{Choice} occurring for the first time in a computation
has to get its own unique identifier.
For instance, if we evaluate the expression \ccode{xor aBool aBool},
the two occurrences of \code{aBool} assign different identifiers
to their \code{Choice} constructor (e.g., \code{1} for the left
and \code{2} for the right \code{aBool} argument) so that this evaluates to
\begin{haskell}
  Choice 1 (Choice 2 False True) (Choice 2 True False)
\end{haskell}
Here we can make different selections for the outer and inner
\code{Choice} constructors so that this non-deterministic result represents
four values.

To summarize, our implementation is based on the following principles:
\begin{enumerate}
\item Each non-deterministic choice is represented by a \code{Choice}
constructor with a unique identifier.
\item When matching a \code{Choice} constructor, the choice is
moved to the result of this operation with the same identifier,
i.e., a non-deterministic argument
yields non-deterministic results for each of the argument's values.
\item Each choice introduced in a computation is supplied with its own unique
identifier.
\end{enumerate}
The latter principle requires the creation of fresh identifiers
during a computation---a non-trivial problem in functional languages.
One possibility is the use of a global counter
that is accessed by unsafe features whenever a new identifier
is required. Unfortunately, unsafe features inhibit the use
of optimization techniques developed for purely functional programs
and make the application of advanced evaluation and search
strategies (e.g., parallel strategies) more complex.
Therefore, we avoid unsafe features in our implementation.
Instead, we thread some global information
through our program in order to supply fresh references
at any point of a computation.
For this purpose, we assume a type \code{IDSupply}
with operations
\begin{haskell}
  initSupply  :: IO IDSupply
  thisID      :: IDSupply -> ID
  leftSupply  :: IDSupply -> IDSupply
  rightSupply :: IDSupply -> IDSupply
\end{haskell}
and add a new argument of type \code{IDSupply} to each operation
of the source program, i.e., a Curry operation of type
\begin{curry}
  f :: $\tau_1$ ->$~\cdots~$->$~\tau_n~$->$~\tau$
\end{curry}
is translated into a Haskell function of type
\begin{haskell}
  f :: $\tau_1$ ->$~\cdots~$->$~\tau_n~$-> IDSupply ->$~\tau$
\end{haskell}
%
Conceptually, one can consider \code{IDSupply}
as an infinite set of identifiers that is created at the beginning
of an evaluation by the operation \code{initSupply}.
The operation \code{thisID} takes some identifier from this set, and
\code{leftSupply} and \code{rightSupply} split this set
into two disjoint subsets without the identifier
obtained by \code{thisID}.
The split operations \code{leftSupply} and \code{rightSupply}
are used when an operation calls two\footnote{The
extension to more than two is straightforward.}
other operations in the right-hand side of a rule. In this case,
the called operations must be supplied with their individual disjoint
identifier supplies. For instance, the operation \code{main}
defined by
\label{sec:xor-main}
\begin{curry}
  main :: Bool
  main = xorSelf aBool
\end{curry}
is translated into
\begin{haskell}
  main :: IDSupply -> Bool
  main s = xorSelf (aBool (leftSupply s)) (rightSupply s)
\end{haskell}
Any choice in the right-hand side of a rule gets its own identifier
by the operation \code{thisID}, as in
\begin{haskell}
  aBool s = Choice (thisID s) True False
\end{haskell}
The type \code{IDSupply} can be implemented in various ways.
The simplest implementation uses unbounded integers:
\begin{haskell}
  type IDSupply = Integer
  initSupply    = return 1
  thisID      n = n
  leftSupply  n = 2*n
  rightSupply n = 2*n+1
\end{haskell}
There are other more sophisticated implementations available
\cite{AugustssonRittriSynek94}.
Actually, our compilation system is parameterized over different
implementations of \code{IDSupply}
in order to perform some experiments and choose
the most appropiate for a given application.
Each implementation must ensure that, if $s$ is a value of type
\code{IDSupply}, then \code{thisID($o_1$($\ldots$($o_n$ $s$)$\ldots$))}
and \code{thisID($o'_1$($\ldots$($o'_m$ $s$)$\ldots$))}
are different identifiers provided that
$o_i,o'_j \in \{\code{leftSupply},\code{rightSupply}\}$
and $o_1\cdots o_n \neq o'_1\cdots o'_m$.

@@

So far, our generated operations compute all the non-deterministic
values of an expression represented by a structure containing
\code{Choice} constructors. In order to extract the various
values from this structure, we have to define operations
that compute all possible choices in some order where the choice identifiers
are taken into account.
To provide a common interface for such operations, we introduce
a data type to represent the general outcome of a computation,
\begin{haskell}
  data Try a = Val a | Choice ID a a | Fail
\end{haskell}
together with an auxiliary operation:\footnote{Note that the
operation \code{try} is not really polymorphic but overloaded
for each data type and, therefore, defined in instances of some type class.}
\begin{haskell}
  try :: a -> Try a
  try (Choice i x y) = Choice i x y
  try Fail           = Fail
  try x              = Val x
\end{haskell}
In order to take the identity of choices into account when extracting values,
one has to remember which choice (e.g., left or right branch)
has been made for some particular choice.
Therefore, we introduce the type
\begin{haskell}
  data Choice = NoChoice | ChooseLeft | ChooseRight
\end{haskell}
where \code{NoChoice} represents the fact that a choice has not yet been made.
Furthermore, we need operations to lookup the current choice
for a given identifier or change its choice:
\begin{haskell}
  lookupChoice :: ID -> IO Choice
  setChoice    :: ID -> Choice -> IO ()
\end{haskell}
In Haskell, there are different possibilities to implement a mapping
from choice identifiers to some value of type \code{Choice}.
Our implementation supports various options together with different
implementations of \code{IDSupply}.
For instance, a simple but efficient implementation can be obtained by
using updatable values, i.e., the Haskell type \code{IORef}.
In this case, choice identifiers are memory cells instead of integers:
\begin{haskell}
  newtype ID = ID (IORef Choice)
\end{haskell}
Consequently, the implementation of \code{IDSupply} requires
an infinite set of memory cells which can represented as a tree structure:
\begin{haskell}
  data IDSupply = IDSupply ID IDSupply IDSupply
  thisID      (IDSupply r _ _) = r
  leftSupply  (IDSupply _ s _) = s
  rightSupply (IDSupply _ _ s) = s
  \end{haskell}
The infinite tree of memory cells (with initial value \code{NoChoice})
can be constructed as follows,
where \code{unsafeInterleaveIO} is used to construct
the tree on demand:
\begin{haskell}
  initSupply = getIDTree$\listline$
  getIDTree = do s1 <- unsafeInterleaveIO getIDTree
                 s2 <- unsafeInterleaveIO getIDTree
                 r  <- unsafeInterleaveIO (newIORef NoChoice)
                 return (IDSupply (ID r) s1 s2)
  \end{haskell}
Using memory cells,
the implementation of the lookup and set operations is straightforward:
\begin{haskell}
  lookupChoice (ID ref) = readIORef ref
  setChoice (ID ref) c = writeIORef ref c
\end{haskell}
Now we can print all values contained in a choice structure
in a depth-first manner by the following operation:
\label{sec:printValsDFS}
\begin{haskell}
  printValsDFS :: Try a -> IO ()
  printValsDFS (Val v)          = print v
  printValsDFS Fail             = return ()
  printValsDFS (Choice i x1 x2) = lookupChoice i >>= choose
   where
    choose ChooseLeft  = printValsDFS (try x1)
    choose ChooseRight = printValsDFS (try x2)
    choose NoChoice    = do newChoice ChooseLeft  x1
                            newChoice ChooseRight x2

@@

In addition to the different search strategies to evaluate
top-level expressions, our system also contains a primitive operation
\begin{curry}
 searchTree :: a -> IO (SearchTree a)
\end{curry}
to translate the search space caused by the evaluation of its argument
into a tree structure of the form
\begin{curry}
 data SearchTree a = Value a | Fail | Or (SearchTree a) (SearchTree a)
\end{curry}
With this primitive,
the programmer can define its own search strategy or
collect all non-deterministic values into a list structure
for further processing \cite{BrasselHuch07}.

@@

In the introduction we already discussed the various
efforts to implement functional logic languages,
like the construction of abstract machines
\cite{AntoyHanusLiuTolmach05,HanusSadre99JFLP,Lux99FLOPS} and
the compilation into Prolog \cite{AntoyHanus00FROCOS}
or Haskell \cite{BrasselFischerHanusReck11,BrasselHuch07,BrasselHuch09}.
Our benchmarks show that an efficient implementation
by compiling into
a functional language depends on carefully handling
the sharing of non-deterministic choices. For instance, our previous
implementation \cite{BrasselFischerHanusReck11}, where sharing
is explicitly managed by the monadic techniques proposed
in \cite{FischerKiselyovShan09}, has not satisfied the expectations
that came from the benchmarks reported in \cite{FischerKiselyovShan09}.
Due to these experiences, in our new compiler we use the
compilation scheme initially proposed in \cite{BrasselFischer08IFL}
which produces much faster code, as shown in our benchmarks.
